---
title: "Using Shiny for Big Data Analysis"
author: "Matthew Hirschey and Akshay Bareja"
date: "8/31/2020"
output: 
  ioslides_presentation:
    incremental: true
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(shiny)
df <- read_csv("df.csv")
genes <- pull(df, Gene.symbol)
```

## What is Shiny? {.build}

- Shiny is an R package that allows you to make interactive apps using R  
- These apps can be run on your computer, hosted as stand-alone apps, or can be embedded in a webpage  
- Think of this as R's version of Tableau  

## What kind of apps can you make with Shiny?

<iframe src="https://shiny.rstudio.com/gallery/#demos" width="100%" height="90%"></iframe>

## How does a Shiny App work? {.build}

- A Shiny app is made up of three components - a user interface section, a server section, and a function to launch the app
- When a user makes some change or enters some input, the app produces the required output (like a graph or table)

## Let's build an app! {.build}

- Using the same microarray dataset we looked at in a previous meetup, let's make an app that plots individual expression values for an input gene
- The user selects a particular gene from a dropdown menu, and the app plots the corresponding graph

##
```{r, echo = FALSE}
shinyApp(
ui <- fluidPage(
    titlePanel("Outlier Detector"),
    sidebarLayout(
        sidebarPanel(
            selectInput("gene", "Gene:", 
                        choices = genes)
        ),
        mainPanel(
           plotOutput("plot")
        )
    )
),

server <- function(input, output) {
    
    output$plot <- renderPlot({
       df %>% 
            filter(Gene.symbol == input$gene) %>% 
            pivot_longer(cols = contains("GSM"),
                         names_to = "Sample",
                         values_to = "Expression") %>% 
            ggplot(aes(Sample, Expression)) +
            geom_point(col = "darkblue", size = 5) +
            theme_bw()
    })
}
)
```

## UI code

```{r ui code, eval = FALSE, echo = TRUE, warning = FALSE, message = FALSE}
library(shiny)
library(tidyverse)

df <- read_csv("df.csv")
genes <- pull(df, Gene.symbol)

ui <- fluidPage(
    titlePanel("Outlier Detector"),
    sidebarLayout(
        sidebarPanel(
            selectInput("gene", "Gene:", 
                        choices = genes)
             # `gene` "smuggles" the selected gene name to the server side
        ),
        mainPanel(
           plotOutput("plot")
        )
    )
)
```

## Server code
```{r server code, eval = FALSE, echo = TRUE}
server <- function(input, output) {

    output$plot <- renderPlot({
       df %>% 
            filter(Gene.symbol == input$gene) %>% 
            pivot_longer(cols = contains("GSM"),
                         names_to = "Sample",
                         values_to = "Expression") %>% 
            ggplot(aes(Sample, Expression)) +
            geom_point(col = "darkblue", size = 5) +
            theme_bw()
       # The output of this code gets assigned to `plot` which is 
       # sent back to the UI side
    })
}
```

## Run the app
```{r run the app, eval = FALSE, echo = TRUE}
shinyApp(ui = ui, server = server)
```

## Adding another reactive element is very simple

##
```{r, echo = FALSE}
shinyApp(
ui <- fluidPage(
    titlePanel("Outlier Detector"),
    sidebarLayout(
        sidebarPanel(
            selectInput("gene", "Gene:", 
                        choices = genes)
        ),
        mainPanel(
           textOutput("text"),
           plotOutput("plot")
        )
    )
),

server <- function(input, output) {
    
    output$plot <- renderPlot({
       df %>% 
            filter(Gene.symbol == input$gene) %>% 
            pivot_longer(cols = contains("GSM"),
                         names_to = "Sample",
                         values_to = "Expression") %>% 
            ggplot(aes(Sample, Expression)) +
            geom_point(col = "darkblue", size = 5) +
            theme_bw()
    })
    
    output$text <- renderText({
      paste0("Expression levels for gene: ", input$gene)
    })
}
)
```

## UI code
```{r, eval = FALSE, echo = TRUE, warning = FALSE, message = FALSE}
library(shiny)
library(tidyverse)

df <- read_csv("df.csv")
genes <- pull(df, Gene.symbol)

ui <- fluidPage(
    titlePanel("Outlier Detector"),
    sidebarLayout(
        sidebarPanel(
            selectInput("gene", "Gene:", 
                        choices = genes)
        ),
        mainPanel(
           textOutput("text"), # new code
           plotOutput("plot") 
        )
    )
)
```

## Server code
```{r, eval = FALSE, echo = TRUE}
server <- function(input, output) {
    
    output$plot <- renderPlot({
       df %>% 
            filter(Gene.symbol == input$gene) %>% 
            pivot_longer(cols = contains("GSM"),
                         names_to = "Sample",
                         values_to = "Expression") %>% 
            ggplot(aes(Sample, Expression)) +
            geom_point(col = "darkblue", size = 5) +
            theme_bw()
    })
    
    output$text <- renderText({ # new code
      paste0("Expression levels for gene:", input$gene) # new code
    }) # new code
}
```

## To run app, open `app.R` in RStudio and click `Run App`

## What is reactivity?

<iframe src="https://shiny.rstudio.com/articles/understanding-reactivity.html" width="100%" height="100%"></iframe>

## Using Shiny apps in publications

https://twc-stanford.shinyapps.io/aging_plasma_proteome/" 
